### 2011-11-19 doc/theses/mikula/proposed_algorithm.tex
Master thesis: In progress.

\begin{itemize}
\item XML Schema built-in atomic types \todo{link do appendixu}.
\item Type $xs:untyped$.
\item Type representing any item. \todo{potrebujeme?}
\item Type representing an element, attribute, item, node, text node, comment, processing instruction, document node.
\item Type representing a set of elements or attributes selected by a certain path expression. The path expression is included in this type.
\end{itemize}

\todo[inline]{Akeho typu je napr //(elementA * 2)?}

@@

\todo[inline]{Popisat co robit a uviest priklad, kedy typ nevieme zistit. Napr porovnavanie hodnot elemetov vybranych z dat.}

This process can be divided into three substeps. Determination of return types of functions, determination of types of global variables, and finally determination of types of expressions.

@@

Instead, the syntax tree can be searched just once, before the analysis of expressions, and return types of all functions found will be stored within an associative array. On request, this array will retrieve the return type of a specified function.

@@

\todo[inline]{Udrzat si referencie na definicie funkcii pre dalsie pouzitie.}

@@

\STATE $type = null$
\IF{$exprNode$ represents a literal value}
     \STATE $type =$ particular literal value type, exactly-once cardinality
\ELSIF{$exprNode$ represents a call of function $f$}
     \STATE $type = getFunctionReturnType(f)$
\ELSIF{$exprNode$ represents variable reference $var$}
     \STATE $type = getVariableType(contextVarTypes, var)$
\ELSIF{$exprNode$ represents path expression $path$}
     \STATE $type = TODO$
\ELSIF{$exprNode$ represents operator $op$}
     \STATE TODO
\ELSIF{$exprNode$ represents $FLWOR$ construct $flwor$}
     \STATE  $type =$ type of an expression in the return clause of $flwor$
\ENDIF

@@

The algorithm iterates through the variable declaration nodes from the prolog and if the variable is explicitly assigned with its type than the type is noted. Otherwise, an attempt of the type deduction of the expression is made. The deduction of the type from the expression is presented in \ref{ALG_func_determineExpressionType}. On input, it takes an expression node and types of local variables that are valid in the expression. The presented algorithm is just shortened illustration since the complete version is too long to be presented in this text. Nevertheless, the principle is straightforward. Depending on the class of the expression, its type can be either determined directly or it depends on its subexpressions. \todo{link na mozne triedy vyrazov.}

@@

How are the types inferred from the expression types? We do not exploit all expressions. Only expressions of a particular type are exploited. Specifically, an expression has to contain a subexpression representing a certain element or attribute or set of elements or attributes. This set is denoted $S$ in the following text. In other words, the type of the subexpression has to contain a path expression that selects the elements or attributes. This subexpression is denoted $E$. Another requirement is the expression has to be either a function call or an arithmetic operation \todo{Nejake ine ako napr type constructor, cast as? Princip je stale rovnaky}.

>>>>

\begin{itemize}
\item XML Schema built-in atomic types \todo{link do appendixu}.
\item Type $xs:untyped$.
\item Type representing any item. \todo{je potrebne?}
\item Type representing an element, attribute, item, node, text node, comment, processing instruction, document node.
\item Type representing a set of elements or attributes selected by a certain path expression. The path expression is included in this type. Let this type be identified as $pathType$
\end{itemize}

\todo[inline]{Akeho typu je napr path expression /root/(elementA * 2)?}

@@

If we cannot determine the type of an expression we express that using $null$. Example of such case is path expression $/rootElement/(elementA * 2)$. \todo{Nejak dopisat}

The analysis of expression types can be divided into three substeps. Determination of return types of functions, determination of types of global variables, and finally determination of types of expressions.

@@

Instead, the syntax tree can be searched just once, before the analysis of expressions, and return types of all functions found are stored.

@@

Later phases of the algorithm may need to process the function declarations. Therefore, we also store references to the entire function declaration nodes.

@@

\STATE $type = null$
\IF{$exprNode$ represents a literal value}
     \STATE $type =$ particular literal value type, exactly-once cardinality
\ELSIF{$exprNode$ represents a call of function $f$}
     \STATE $type = getFunctionReturnType(f)$
\ELSIF{$exprNode$ represents variable reference $var$}
     \STATE $type = getVariableType(contextVarTypes, var)$
\ELSIF{$exprNode$ represents path expression $path$}
     \IF{$path$ selects an element set}
         \STATE $type =$ element type, zero-or-more cardianlity, $path$ included
     \ELSIF{$path$ selects an attribute set}
         \STATE $type =$ attribute type, zero-or-more cardianlity, $path$ included
     \ELSE
         \STATE $type = null$
     \ENDIF
\ELSIF{$exprNode$ represents operator $op$}
     \IF{$op$ is a comparison operator}
         \STATE $type = xs:boolean$
     \ELSIF{$op$ is an additive operator $AND$ one of the operands is of a numeric type}
         \IF{both operands are of numeric types}
             \STATE $type =$ the numberic type which is more or equally general than the other
         \ELSE
             \STATE $type =$ type of the numeric operand
         \ENDIF
     \ENDIF
\ELSIF{$exprNode$ represents $FLWOR$ construct $flwor$}
     \STATE  $type =$ type of an expression in the return clause of $flwor$
\ENDIF

@@

The algorithm iterates through the variable declaration nodes from the prolog and if the variable is explicitly assigned with its type than the type is noted. Otherwise, an attempt of the type deduction of the expression is made. The deduction of the type from the expression is presented in \ref{ALG_func_determineExpressionType}. On input, it takes an expression node and types of local variables that are valid in the expression (also called context of variables or variable context). The presented algorithm is just shortened illustration since the complete version is too long to be presented in this text. Nevertheless, the principle is straightforward. Depending on the class of the expression, its type can be either determined directly or it depends on its subexpressions. \todo{link na mozne triedy vyrazov.}

@@

How are the types inferred from the expression types? We do not exploit all expressions. Only expressions of a particular type are exploited. Specifically, an expression has to contain some $pathType$ subexpression $E$ (expression representing a certain element or attribute or set of elements or attributes). In the following text, the set represented by expression $E$ is denoted $S$. Another requirement is the expression has to be either a function call or an arithmetic operation \todo{Nejake ine ako napr type constructor, cast as? Princip je stale rovnaky}.
